---
title: "Tidying Data"
author: "Paloma Cartwright"
date: "16/08/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

```{r}

us_rent_income_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable,
              values_from = estimate)

```

## Practice some tidyverse functions for data wrangling 

- `filter()`
- `select()`
- `mutate()`
- `case_when()`
- `group_by() %>% summarize()`
- piping through things in sequence 

### dplyr::filter() to keep or exclude rowas based on our conditions 

1. Exact match for a single condition 
Only keep observaitons where the island is Biscoe
```{r}
 
ex_1 <- filter(.data = penguins, 
               island == "Biscoe")

```

2. "and' statement within filter 
Only keep observations if the penguin species is Chinstrap AND the year is 2008 

```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)
  # the two lines below do the exact same thing 
  # filter(species == "Chinstrap" & year == 2008)
  # filter(species == "Chinstrap") %>% filter(year == 2008)
```

3. Exclude Gentoos 

Keep observations is the penguin species is NOT "Gentoo"

```{r}
ex_3 <- penguins %>% 
  filter(!species == "Gentoo")
```

Exercise: 
Create a subset from `penguins` that only contains observations for Adelie penguins with flipper length >= 185mm. 

```{r}

exercise_1 <- penguins %>% 
  filter(flipper_length_mm >= 185, species == "Adelie")
```

4. The %in% operator 

Use the `%in%` operator to look for *any* matches within a vector and return TRUE if the value does exist for our row. 
We want to keep observations where the island is Dream or Torgersen. 

```{r}
ex_4 <- penguins %>%  
  filter((island %in% c("Dream", "Torgersen")))
```

5. Keep a single column by name 

Only want to keep the speceis column 

```{r}
ex_5 <- penguins %>% select(species)


# Reminder, rewrite this without the pipe operator 

# ex_5 <- select(.data = penguins, species)
```

6. Let's keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year

```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

7. Choose or exclude columns based on position 

```{r}
ex_7 <- penguins %>% 
  select(-4)
```

8. A piped sequence 

Write a piped sequence where you first only keep columns for species, body_mass_g, and island and second, filter to only include observations if body mass is greater than 3600 

```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

### mutate() to add or update existing columns, while keeping the ones already there

9. Add a column that has body mass converted to kg

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000, #) %>% 
  # mutate(
  flipper_length_m = flipper_length_mm / 1000)
```

10. Mutate to update an existing column

```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", 
                              replacement = "yeah"))
```

11. Update the class of the bill depth column to caracter 

```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))

```


### case_when(): your favorite if-else 

12. Create a new column that contains "small" is penguin flipper length < 185, "medium" otherwise

```{r}
ex_12 <- penguins %>%  
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: Starting from penguins, create a new column called "bird_bill_bin" that contains: 
- if bill length < 38 mm: "short bill" 
- if bill length >= 38mm & < 43: "medium bill" 
- if bill lenght >= 43: "long bill" 

```{r}
ex_13 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill", 
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill", 
    bill_length_mm >= 43 ~ "long bill" 
    # TRUE ~ "long bill" same thing as above but this includes na values 
  ))
```

### group_by() %>% summarize() to find summary values by group and return them in a nice summary table 

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```
14. group_by() to say there are groups within the speceis column 
```{r}
ex_14 <- penguins %>% 
  group_by(species, island, year, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, 
                                       na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, 
                                   na.rm = TRUE))

ex_15 <- penguins %>% 
  group_by(species) %>% 
  mutate(mean_flipper_length2 = mean(flipper_length_mm, na.rm = TRUE))
```

## Practice: 

In a single pipe sequence do the following: 
- Only keep penguins that have body mass > 3500 
- only keep columns for species, flipper_length_mm and sex 
_ Create a summary table that contains the mean, sd, max and min value of flipper length by species and sex.

```{r}
ex_16 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = 
              mean(flipper_length_mm, na.rm = TRUE), 
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE), 
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE), 
            min_flip = min(flipper_length_mm, na.rm = TRUE)) %>% 
  ungroup() 
  
```

